Species turnover reveals hidden effects of decreasing Nitrogen deposition in mountain hay meadows

### 2018-06-29 Manuscript.Rmd
Update Manuscript.Rmd

Observational studies along a gradient of Nitrogen deposition often infer how the spatial variation in species richness is related to Nitrogen deposition [@Stevens2010gradient; chapter 4 in @deVries2015]. They assume that the spatial variation in species richness (or other metrics of community composition) arose because over time some areas lost more species than others because they chronically experienced higher Nitrogen deposition. Although there is evidence supporting the use of such a 'space for time substitution' for detecting the effects of Nitrogen deposition on plant diversity [@Stevens2010], they can not replace studies that relate temporal patterns in species with Nitrogen deposition [@deSchrijver2011]. There are only a limited number of studies inferring temporal trends of plant species diversity related to varying amounts of Nitrogen deposition in extant communities [@Clark2008; @Storkey2015; @Stevens2016]. In a recent study, we used the mountain hay meadow data from a single survey only and estimated the critical load along the Nitrogen deposition gradient at which species richness of oligotrophic species richness start to decrease with increasing Nitrogen deposition [@Roth2017]. Using this spatial variation in species richness and nitrogen deposition we estimated a critical load for mountain hay meadows of 13.1 kg ha$^{-1}$ yr$^{-1}$. In the current study we estimated the rate of Nitrogen deposition at which local survival probability or colonization probability was equal for oligotrophic and eutrophic species. This rate could also be interpreted as a critical load. With this temporal approach we obtained very similar results as the spatial approach. Our results are thus validating the space for time approach, at least for Swiss mountain hay meadows.

However, the results of Fig. \@ref(fig:figconsequences) also shows that the results of spatial comparisons must be interpreted carefully. When we used the same model to infer how oligotrophic species richness is decreasing along the Nitrogen deposition gradient at different time points, except for the first two study periods the decrease became steeper (more negative) over time. Our first interpretation was that the Nitrogen deposition effect became stronger over time. This contradicted our prediction that the Nitrogen deposition should become weaker over time since Nitrogen deposition decreased. Only then we realized that species turnover was highest in low deposition sites (Tab. \@ref(tab:turnovertab)) where the colonizing species were species with below average indicator values for Nutrients. It seems that the decrease in Nitrogen deposition resulted in oligotrophic species replacing eutrophic species especially at sites with low deposition, thus leading to a steeper spatial decline in oligotrophic species richness from sites with low to sites with high Nitrogen deposition. The increasingly steeper decline over time in the eutrophic species richness along the spatial Nitrogen deposition gradient is thus a sign of the recovery of plant communities in low deposition sites rather than a sign that the negative Nitrogen deposition effects became stronger over time.

# Conclusions {-}
  Comparing the indicator values of colonizing and disappearing species with random expectation we found that oligotrophic species are currently more likely to colonize mountain hay meadows than oligotrophic species, which might be the results of the recently decreasing atmospheric Nitrogen deposition. However, our results also caution that re-transferring mountain hay meadows to oligotrophic species communities after high nitrogen deposition might take much longer than transferring an oligotrophic into an eutrophic community. This is because eutrophic species have high local survival probabilities also when Nitrogen deposition becomes lower again. All these results were, however, hidden when only inferring the temporal trends of species communities. Our study thus adds to the understanding of contemporary biodiversity change that is currently not well understood [@Magurran2018], and it supports the notion of @Hillebrand2018 that inferring species turnover will generate a far more reliable understanding of the biotic response to changing environments than solely tracking community composition.

# Acknowledgements {-}
We thank the dedicated and qualified botanists who conducted fieldwork. The Swiss Federal Office for the Environment (FOEN) kindly provided biodiversity monitoring data and topographic data. This work was supported by the FOEN, the Swiss National Science Foundation (grant no. 31003A_156294), the Swiss Association Pro Petite Camargue Alsacienne, the Fondation de bienfaisance Jeanne Lovioz, and the MAVA Foundation.

>>>>

Observational studies along a gradient of Nitrogen deposition often conclude how the spatial variation in species richness is related to Nitrogen deposition [@Stevens2010gradient; chapter 4 in @deVries2015]. They assume that the spatial variation in species richness (or other metrics of community composition) arose because of unequal species loss of different areas over time resulting from elevated N-deposition chronically experienced by some areas. Although there is evidence supporting the pertinency of such a 'space for time substitution' for detecting the effects of Nitrogen deposition on plant diversity [@Stevens2010], this approach can not replace studies that relate temporal patterns in species composition with Nitrogen deposition [@deSchrijver2011]. There are only a limited number of studies directly relating temporal trends of plant species diversity to varying amounts of Nitrogen deposition in extant communities [@Clark2008; @Storkey2015; @Stevens2016]. In a recent study, we used the mountain hay meadow data from a single survey only and estimated the critical load along the Nitrogen deposition gradient at which species richness of oligotrophic species richness start to decrease with increasing Nitrogen deposition [@Roth2017]. Using this spatial variation in species richness and nitrogen deposition we estimated a critical load for mountain hay meadows of 13.1 kg ha$^{-1}$ yr$^{-1}$. In the current study we estimated the rate of Nitrogen deposition at which local survival probability or colonization probability was equal for oligotrophic and eutrophic species. This rate could also be interpreted as a critical load. With this temporal approach we obtained very similar results as the spatial approach. Our results are thus validating the space for time approach, at least for Swiss mountain hay meadows.

However, the results of Fig. \@ref(fig:figconsequences) also show that the results of spatial comparisons must be interpreted carefully. When we applied the same model as in Roth et al. (2017) for different study periods to infer how oligotrophic species richness is decreasing along the Nitrogen deposition gradient, the relationship showed to vary between study periods. The decrease gradually became steeper (more negative) over time, except for the first two study periods. Our first interpretation was that the Nitrogen deposition effect became stronger over time. This was against our prediction that the effect of N-deposition should become weaker over time since Nitrogen deposition actually decreased in the course of the study period. Then we realized that species turnover was highest at low N-deposition sites (Tab. \@ref(tab:turnovertab)). At low N-deposition rates colonizing species have below average indicator values for nutrients. It seems that the decrease in Nitrogen deposition resulted in oligotrophic species replacing eutrophic species especially at sites with low N-deposition. This explains why the decline in oligotrophic species richness inferred from spatial patterns of species richness and N-deposition is becoming steeper over time. This may be interpreted as evidence that plant communities are recovering at least at low deposition sites rather than inferring that the negative N-deposition effects have become stronger over time.

# Conclusions {-}
  Comparing the indicator values of colonizing and disappearing species with random expectation we found that oligotrophic species are currently more likely to colonize mountain hay meadows than eutrophic species, which might be the result of decreasing atmospheric Nitrogen deposition recently observed. However, our results also indicate that re-transferring mountain hay meadows to oligotrophic species communities after high nitrogen deposition might take much longer than transferring an oligotrophic into an eutrophic community. This is because eutrophic species have high local survival probabilities also when Nitrogen deposition becomes lower again. However, these results are hidden if only the temporal trends of species composition are analysed instead of species turnover, which can better reveal ongoing processes in this case. Our study thus adds to the understanding of contemporary biodiversity change where some key findings are still missing [@Magurran2018], and it supports the notion of @Hillebrand2018 that consicering species turnover will generate far more reliable view of the biotic response to changing environments than solely tracking community composition.

# Acknowledgements {-}
We thank the dedicated and qualified botanists who conducted fieldwork for the Swiss Biodiversity Monitoring BDM. The Swiss Federal Office for the Environment (FOEN) kindly provided biodiversity monitoring data and topographic data. This work was supported by the FOEN, the Swiss National Science Foundation (grant no. 31003A_156294), the Swiss Association Pro Petite Camargue Alsacienne, the Fondation de bienfaisance Jeanne Lovioz, and the MAVA Foundation.
